Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-02-15 paper/rebuttal/rebuttal.tex

Kind Regards,

Dakota, Jisung, Sadamori, Rodrigo, Woo-Sung, Staša, and YY

\clearpage
\tableofcontents
\clearpage

\section{Theoretical justification of \textit{word2vec} and the Gravity Model}
\label{sec:theory}
Here, we answer a question: Why does the skip-gram model with negative sampling (SGNS), otherwise known as \textit{word2vec}, work so well to model human mobility?
We demonstrate a fundamental connection between the SGNS model and the Gravity Model of Mobility, namely, that SGNS implicitly learns gravity-like relationships between locations.
This connection provides necessary theoretical justification to our choice of \textit{word2vec}, and constitutes a major breakthrough of our understanding of neural embeddings.
In addition, this motivates the use of SGNS for future study of human mobility.

@@

Taken altogether, the {\it word2vec} model with the negative sampling predicts a flow in the same form as in the gravity model:
\begin{align}
	\hat{T_{ij}}=  C P(i) P(j)  \exp(\bm{v}_j \cdot \bm{v}_{i}), \label{eq:flow_w2v_ng0}
\end{align}
where $C$ is a positive constant.

In other words, \emph{word2vec with the negative sampling is a gravity model}, with the mass given by the location's frequency $P(i)$, and the distance measured by their dot similarities.
While the gravity model generates the flow from the given mass and locations, {\it word2vec} finds the mass and locations that best explain the given trajectories.
Therefore, the embedding distance is inherently tied to the flow, and hence, has greater predictive power than the geographic distance.

\subsection{Gap between the mathematical analysis and numerical simulations}

In the mathematical analysis, we assumed that the embedding vectors are optimal in terms of the objective of {\it word2vec}, which, however, is not true in our numerical simulations.
This violation of assumption leads to two substantial consequences for embedding vectors.
First, although there exist optimal in-vector and out-vector that are identical, the in-vector and out-vector can be different in practice because they are sub-optimal.
Second, while dot similarity is a formal distance measure in terms of its connection to the gravity model, our simulations show that dot similarity did not predict mobility flow better than cosine similarity (Fig.~\ref{fig:supp:distancemetrics}). It has been widely observed that cosine similarity is generally better than dot similarity in many applications of word embedding~\autocite{levy2015improving}. Therefore, the superiority of cosine similarity is not a problem specific to, but instead a general feature of word embedding.
Nevertheless, dot similarity still outperformed geographic distance and some baselines (Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}). What makes cosine similarity superior to dot similarity warrants the future work.

### 2021-02-17 paper/current/main.tex
update main paper

A vector-space embedding of locations (airports, accommodations, and organizations) is learned by using trajectories as input to the standard with skip-gram negative sampling, or \textit{word2vec} neural-network architecture (see Methods).
This neural embedding model, originally designed for learning language models~\autocite{mikolov2013word2vec}, has been making breakthroughs by revealing novel insights into texts~\autocite{tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered}, networks~\autocite{grover2016node2vec, linzhuo2020hyperbolic} and trajectories\autocite{liu2016exploring, feng2017poi2vec, yao2018representing, cao2019habit2vec, crivellari2019motion, solomon2018predict}.
It works under the notion that a good representation should facilitate prediction, and so learns a mapping of target locations (words) that best predict locations that co-appear in the same trajectory (sentence).
The model is also computationally efficient, robust to noise, and can encode relations between entities as geometric relationships in the vector space~\autocite{levy2014neural, nakandala2017gendered, kozlowski2018geometry, an2018semaxis}.
As a result, each location is encoded into a single vector representation, and vectors relate to one another based on the likelihood of locations appearing adjacent to one another in the same trajectory.

@@

A central assumption in our mathematical analysis is that the embedding vectors are optimum in terms of the {\it word2vec}'s objective function.
However, this premise is not practical because the optimization algorithm often finds sub-optimal embeddings due to its stochasticity and being trapped in a local optimum.
This has two consequences.
First, the in- and out-vectors can be different.
Second, dot similarity should in principle predict the mobility flow better than cosine similarity because dot similarity includes cosine similarity as its special case.
However, our simulations show that it is the cosine similarity that better predicted the mobility flow~[reference to a figure].
This phenomenon---cosine similarity outperforms dot similarity---has been observed in many applications of word embedding~\cite{add citations}.
Nevertheless, among the 7(?) measures tested in our simulations, dot similarity is the second-best predictor of mobility flow following cosine similarity.
Furthermore, unlike cosine similarity, dot similarity has a mathematical connection to the gravity model.

\todo{add stuff about cosine distance performing better, SGD performing better}

@@

There is an alternative implementation of \textit{word2vec} based on a matrix factorization~\cite{levy2014neural,Qiu2018}, which optimizes the same objective function as for \texttt{gensim} but in a different way.
The matrix factorization approach generates embedding vectors such that the dot similarity for locations $i$ and $j$ is as close as $q_{ij}= \log(T_{ij} / s_i s_j) + \text{const}$.
An issue is that the generated embedding vectors can be largely affected by small flows $T_{ij}$ in the data.
For example, if there is no flow between locations $i$ and $j$ (\ie, $T_{ij} = 0$), then $q_{ij} = -\infty$, pushing $i$ and $j$ to the infinite horizon of the embedding space.
To remedy this problem, one can add a small amount of flow between such locations or exclude them from the embedding~\cite{levy2014neural,Qiu2018}.
However, this may change the data to which we fit the model, and small flows may still have profound impact on the embedding space.
Therefore, we adopt the \textt{gensim} implementation of the \textit{word2vec}.
Our simulations show that, in the prediction task of mobility flow, \texttt{gensim}'s word2vec performs better than the matrix factorization approach in xxx out of the xxx datasets (Fig.~??).

>>>>

A vector-space embedding of locations (airports, accommodations, and organizations) is learned by using trajectories as input to the standard with skip-gram negative sampling, or \textit{word2vec} neural-network architecture (see Methods).
This neural embedding model, originally designed for learning language models~\autocite{mikolov2013word2vec}, has been making breakthroughs by revealing novel insights into texts~\autocite{tshitoyan2019mat2vec, garg2018gender, kozlowski2018geometry, hamilton2016diachronic, le2014doc2vec, nakandala2017gendered}, networks~\autocite{grover2016node2vec, linzhuo2020hyperbolic} and trajectories\autocite{liu2016exploring, feng2017poi2vec, yao2018representing, cao2019habit2vec, crivellari2019motion, solomon2018predict}.
It works under the notion that a good representation should facilitate prediction,  so learns a mapping of words that predicts the target word with surrounding words (context).
The model is also computationally efficient, robust to noise, and can encode relations between entities as geometric relationships in the vector space~\autocite{levy2014neural, nakandala2017gendered, kozlowski2018geometry, an2018semaxis}.
As a result, each location is encoded into a single vector representation, and vectors relate to one another based on the likelihood of locations appearing adjacent to one another in the same trajectory.

@@

A central assumption in our mathematical analysis is that the embedding vectors are optimum in terms of the {\it word2vec}'s objective function.
However, this premise is not practical because the optimization algorithm (e.g. stochastic gradient descent algorithm) often finds sub-optimal embeddings due to its stochasticity and being trapped in a local optimum.
This has two consequences.
First, the in- and out-vectors can be different.
Second, dot similarity should in principle predict the mobility flow better than cosine similarity because dot similarity includes cosine similarity as its special case.
However, our simulations show that it is the cosine similarity that better predicted the mobility flow~(Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
 It has been widely observed that cosine similarity (dot product with $l2$ normalization) is generally better than dot similarity and dot similarity with various normalization methods in capturing a similarity between two words ~\autocite{levy2015improving}.
Nevertheless, among the ten measures tested in our simulations, dot similarity is the second-best predictor of mobility flow following cosine similarity.

@@

There is an alternative implementation of \textit{word2vec} based on a matrix factorization~\cite{levy2014neural,Qiu2018}, which optimizes the same objective function as for \texttt{gensim} but in a different way.
The matrix factorization approach generates embedding vectors such that the dot similarity for locations $i$ and $j$ is as close as $q_{ij}= \log(T_{ij} / s_i s_j) + \text{const}$.
An issue is that the generated embedding vectors can be largely affected by small flows $T_{ij}$ in the data.
For example, if there is no flow between locations $i$ and $j$ (\ie, $T_{ij} = 0$), then $q_{ij} = -\infty$, pushing $i$ and $j$ to the infinite horizon of the embedding space.
To remedy this problem, one can add a small amount of flow between such locations or exclude them from the embedding~\cite{levy2014neural,Qiu2018}.
However, this may change the data to which we fit the model, and small flows may still have profound impact on the embedding space.
Therefore, we adopt the \texttt{gensim} implementation of the \textit{word2vec}.
Our simulations show that, in the prediction task of mobility flow, \texttt{gensim}'s word2vec performs better than the matrix factorization approach in xxx out of the xxx datasets (Fig.~??).
